Theory of surface second-harmonic generation for semiconductors including effects of nonlocal operators

### 2014-11-25 paper-v-0.tex
reverted intro and abstract to previous version

\begin{abstract}
We formulate a theoretical approach of surface second-harmonic generation from 
semiconductor surfaces based on the length gauge. We calculate the nonlinear 
second-order surface susceptibility tensor 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
independent particle approximation, and for the first time in one unique 
formulation we include i) the scissors correction, needed to have the correct 
value for the energy band gap, ii) the derivation for the inclusion of the cut 
function used to extract the surface nonlinear response, and iii) the 
contribution of the nonlocal part of the pseudopotentials that are routinely 
used in \textit{ab initio} band structure calculations. The first and third 
terms are described by spatially nonlocal quantum mechanical operators and 
are fully taken into account in the present formulation. As a test-case for 
this approach, we calculate $\chi^{xxx}(-2\omega;\omega,\omega)$ for a clean 
Si(001)$2\times 1$ reconstructed surface and find that this scheme for 
extracting the surface response through the cut function is exact and readily
implemented. The effects of the scissors correction and the nonlocal part of 
the pseudopotentials are discussed for the first time for surface nonlinear 
optics. The scissors correction non-rigidly shifts the spectrum to higher 
energies by mixing the $1\omega$ and $2\omega$ resonances and exerts a strong 
influence on the line-shape. The effects of the nonlocal part of the 
pseudopotentials keeps the same line-shape of 
$|\chi^{xxx}_{2\times 1}(-2\omega;\omega,\omega)|$ but reduces its value by 
15-20\%. Therefore, our scheme is unprecedented through the inclusion of 
these three contributions and also opens the possibility to study SSHG with 
more versatility and more accuracy than previous approaches. 
\end{abstract}

@@

The paper is organized as follows. In Sec. \ref{theory} we present the 
relevant steps for deriving the surface second-order susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
length gauge formalism. This derivation includes the addition of the terms 
mentioned above that have been absent in previous works. If desired, we can 
use $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ for a 
layer-by-layer analysis. In Sec. \ref{results} we show the results of this 
reformulation with a study of a clean Si(001) surface with a $2 \times 1$ 
reconstruction that proves the validity of our approach with a special 
test-case that has never been implemented before. We compare results from 
before and after adding the different nonlocal contributions. Finally, in 
Sec. \ref{conc} we submit our conclusions.

>>>>

\begin{abstract}
We formulate a theoretical approach of surface second-harmonic generation 
from semiconductor surfaces based on the length gauge. 
Within the independent particle approximation 
the nonlinear second-order surface susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$   
is calculated, including for the first time in one unique formulation
i) the scissors correction, needed to have the correct value of the
energy band gap,
ii) the derivation for the inclusion of the cut function, 
used to extract the surface nonlinear response,
and iii) the contribution of the nonlocal part of the 
pseudopotentials, routinely  used in \textit{ab initio} band structure calculations. 
First and third terms are described by spatially nonlocal quantum
mechanical operators and are fully taken into account in the present formulation. 
As a test-case of the approach we calculate 
$\chi^{xxx}(-2\omega;\omega,\omega)$ for the clean Si(001)$2\times 1$
reconstructed surface and find that our scheme to extract the surface
through the cut function
readily works and is exact. The effects of the scissors correction and
of the nonlocal part of the pseudopotentials are discussed for the
first time in surface nonlinear optics. 
The scissors correction shifts the spectrum to 
higher energies though the shifting is not rigid and mixes the 
$1\omega$ and $2\omega$ resonances, 
and has a strong influence in the line-shape. 
The 
effects of the nonlocal part of the pseudopotentials 
keeps the same line-shape of $|\chi^{xxx}_{2\times 1}(-2\omega;\omega,\omega)|$, but 
reduces its value
by 15-20\%.  
Therefore, the inclusion of the three aforementioned 
contributions is very important and 
makes our scheme unprecedented and opens the possibility to 
study SSHG with more versatility and providing more accurate results 
than previous approaches. 
\end{abstract}

% \begin{abstract}
% We formulate a theoretical approach of surface second-harmonic generation from 
% semiconductor surfaces based on the length gauge. We calculate the nonlinear 
% second-order surface susceptibility tensor 
% $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
% independent particle approximation, and for the first time in one unique 
% formulation we include i) the scissors correction, needed to have the correct 
% value for the energy band gap, ii) the derivation for the inclusion of the cut 
% function used to extract the surface nonlinear response, and iii) the 
% contribution of the nonlocal part of the pseudopotentials that are routinely 
% used in \textit{ab initio} band structure calculations. The first and third 
% terms are described by spatially nonlocal quantum mechanical operators and 
% are fully taken into account in the present formulation. As a test-case for 
% this approach, we calculate $\chi^{xxx}(-2\omega;\omega,\omega)$ for a clean 
% Si(001)$2\times 1$ reconstructed surface and find that this scheme for 
% extracting the surface response through the cut function is exact and readily
% implemented. The effects of the scissors correction and the nonlocal part of 
% the pseudopotentials are discussed for the first time for surface nonlinear 
% optics. The scissors correction non-rigidly shifts the spectrum to higher 
% energies by mixing the $1\omega$ and $2\omega$ resonances and exerts a strong 
% influence on the line-shape. The effects of the nonlocal part of the 
% pseudopotentials keeps the same line-shape of 
% $|\chi^{xxx}_{2\times 1}(-2\omega;\omega,\omega)|$ but reduces its value by 
% 15-20\%. Therefore, our scheme is unprecedented through the inclusion of 
% these three contributions and also opens the possibility to study SSHG with 
% more versatility and more accuracy than previous approaches. 
% \end{abstract}

@@

The paper is organized as follows. In Sec. \ref{theory} we present the 
relevant steps for deriving the surface second-order susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
length gauge formalism. This derivation includes the addition of the terms 
mentioned above that have been absent in previous works. Also, our 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ can be used 
for a layer-by-layer analysis if desired. In Sec. \ref{results} we show the 
results of this reformulation with a study of a clean Si(001) surface with a 
$2 \times 1$ reconstruction, 
proving the correctness of our approach with a special test-case,
never exploited before.
We compare results from before and after adding 
the different nonlocal contributions. Finally, in Sec. \ref{conc} we give 
our conclusions.

% \section{Introduction}\label{intro}

% In recent years surface nonlinear optical spectroscopies, particulary surface 
% second-harmonic generation (SSHG), have evolved as useful nondestructive 
% and noninvasive tools to study surface and interface properties. These 
% properties include atomic structure, phase transitions, adsorption of 
% atoms, and many others.\cite{daumPRL93, mcgilpOE94, meyerPRL95, powerPRL95, 
% godefroyAPL96, hoferAPA96, dadapPRB97, bloembergenAPB99, mcgilpSRL99, 
% suzukiAPB99, mitchellSS01, hughesPRB96, guyotPRB88, downerPSSA01, shenAPB99, 
% shenNAT89, chenPRL81, mendozaPRL98, downerSIA01} Nowadays, SSHG spectroscopy 
% is a crucial tool for research and development in microelectronics,
% \cite{zheltikovLP00} semiconductors, \cite{lupkeSSR99} nanomaterials,
% \cite{salazar-aparicioPRB14} and many more recent areas of great scientific 
% and commercial interest.\cite{cazzanelliNM14} The high surface sensitivity 
% of SSHG spectroscopy is due to the fact that within the dipole approximation 
% the bulk SHG signal of centrosymmetric materials is identically zero. 
% Radiation can occur only at the surface where the inversion symmetry is 
% broken. Much of the foundation of surface science has been built from 
% experiments involving emission or scattering of electrons from surfaces. 
% These require ultrahigh vacuum (UHV) environments and provide no access to 
% buried interfaces. However, SSHG is compatibile with non-UHV conditions and 
% has access to interfaces buried beneath transparent overlayers. Even when 
% applied to surfaces in UHV, the light source and detectors can be aligned and
% used outside the vacuum chamber.

% The usefulness of SSHG could be limited by the lack of microscopic theoretical
% understanding of the nonlinear spectra. The macroscopic phenomenological 
% theory of SSHG, which relates the intensity, phase, and polarization of 
% detected fields to nonlinear and linear susceptibilities at the material 
% interface is now fully developed. However, microscopic theory that relates 
% atomic-level structure to the nonlinear source polarization is still being 
% developed. In this article we put forward an approach to calculate the
% microscopic second-harmonic surface susceptibility that encompasses several
% theoretical features not taken into account before. Within the independent 
% particle approximation (IPA) some frameworks for bulk SHG have been developed 
% to study the nonlinear optical response of bulk materials with varying degrees 
% of sophistication.\cite{butcherPOPS63, aspnesPRB72, sipePRB93, levinePRB94,
% aversaPRB95, hughesPRB96, rashkeevPRB98}

% The most used framework for \textit{ab initio} calculations, Density 
% Functional Theory (DFT) within the Local Density Approximation (LDA),
% \cite{kohnPR65} underestimates the energy band gap of semiconductors. It is 
% well understood that one has to include the many-body interaction to correct 
% for this underestimation of the gap. In this context, the so-called GW 
% approximation\cite{onidaRMP02} is known to correct the band gap of most 
% semiconductors\cite{luceroJPCM12}. However this can be a very expensive
% calculation and thus one uses the much simpler scissors operator scheme.
% \cite{levinePRL89,levinePRL91,delsolePRB93} This allows us to ``open'' the 
% DFT-LDA gap to its correct experimental or GW value even when experiments 
% are not available. In this context, to correct for the underestimation of 
% the energy band gap of semiconductors Nastos et al.\cite{nastosPRB05} used 
% the ``length gauge'' or ``$\mathbf{r}\cdot\mathbf{E}$ gauge'' to show how to 
% correctly include the many-body corrections through the scissors operator in 
% the SH susceptibility. Later, Cabellos \textit{et al}.\cite{cabellosPRB09} 
% elaborated a derivation of the ``velocity gauge'' or 
% ``$\mathbf{A}\cdot\mathbf{v}$'' gauge properly including the scissors operator 
% and proved gauge invariance with respect to the length gauge. From these works 
% it is clear the length gauge is a much better staring point to obtain the 
% surface SH (SSH) susceptibility, as will be elaborated in this article. 
% However, these considerations are only valid for bulk semiconductors.

% Concerning the optical response of surfaces and interfaces, Reining 
% \textit{et al.}\cite{reiningPRB94} introduced the concept of a cut function 
% in order to obtain the surface SH susceptibility tensor. This cut function 
% is required since one usually uses a slab approach when treating semi-infinite 
% surface systems. If the slab is centrosymmetric the susceptibility tensor will 
% be identically zero. The cut function is such that it separates the nonlinear 
% response for the two surfaces of the slab avoiding the destructive 
% interference between them giving a finite value that one identifies with the 
% SSH susceptibility tensor. If the slab is not centrosymmetric the cut function 
% can be used to separate the different signals coming from either surface of 
% the slab. Indeed, one of the main results of this article is to show that the 
% SSH  susceptibility tensor obtained by using the cut function is correctly 
% extracted from the slab. After Reining \textit{et al.},\cite{reiningPRB94}
% Refs.~\onlinecite{mendozaPRL98,arzatePRB01,mendozaPRB01,mejiaPRB02,sanoPRB02}
% followed upon this work and in particular Ref.~\onlinecite{arzatePRB01} went 
% into a  detailed analysis of the different contributions to the SHG spectra 
% of a surface and the nuanced relationship between bulk, surface, interband, 
% intraband, $1\omega$ and $2\omega$ terms, and
% Ref. \onlinecite{mejiaRMF04} developed a layer-by-layer 
% analysis for the nonlinear responses of semiconductor systems, within a 
% tight-binding framework. This model allows for obtaining results from 
% selected regions of a system including the surface. However, in these 
% references the scissors operator is either excluded or incorrectly 
% implemented.

% Finally, DFT-LDA calculations are often based on the use of pseudopotentials.
% As we will discuss later in this article, the presence of the nonlocal part 
% of the pseudopotential introduces corrections to the momentum operator of 
% the electron that have to be included with care into the SSH susceptibility. 
% So far, this contribution has been neglected in the literature.

% Therefore, within the IPA the most complete approach for the calculation of 
% the SSH susceptibility is one which includes i) the scissors correction, 
% ii) the cut function, and iii) the contribution of the nonlocal part of the 
% pseudopotential. Therefore, the goal of this paper is to derive a new 
% expression within the length gauge for the SSH susceptibility tensor 
% $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ 
% that include the aforementioned contributions. The inclusion of these three 
% contributions makes our scheme unprecedented and opens the possibility to 
% study surface SHG with more versatility and providing more accurate results 
% than previous approaches.

% The paper is organized as follows. In Sec. \ref{theory} we present the 
% relevant steps for deriving the surface second-order susceptibility tensor
% $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
% length gauge formalism. This derivation includes the addition of the terms 
% mentioned above that have been absent in previous works. If desired, we can 
% use $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ for a 
% layer-by-layer analysis. In Sec. \ref{results} we show the results of this 
% reformulation with a study of a clean Si(001) surface with a $2 \times 1$ 
% reconstruction that proves the validity of our approach with a special 
% test-case that has never been implemented before. We compare results from 
% before and after adding the different nonlocal contributions. Finally, in 
% Sec. \ref{conc} we submit our conclusions.
